Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation 

### 2025-05-21 manuscript/manuscript.qmd
added supplementary materials and moved eqns there

@@

|   | log_MBR              | MBR_flag   | log_chl              | hplc_flag   |
|----:|:-----------------------|:-----------|:-----------------------|:------------|
| 905 | -0.2545615131431814  | Rrs510     |  0.5837653682849998  | hplc        |
|  91 | -0.3641327253450254  | Rrs510     |  1.0511525224473812  | fluo        |
| 528 | -0.0737170739918758  | Rrs510     |  0.14363923527454328 | fluo        |
| 969 |  0.5813600497877703  | Rrs411     | -0.8272803386505035  | fluo        |
| 552 |  0.7134341855521412  | Rrs443     | -0.9318141382538383  | hplc        |
::: A random 5-row sample from the Pandas dataframe holding the data used in subsequent models. *log_MBR* is the log-transformed Max. Band Ratio (MBR) of sea surface reflectances, where the numerator is Max(Rrs411, Rrs443, Rrs489, Rrs510) and the denominator is computed from the sum of Rrs555 and Rrs670. *MBR_flag* indicates which of the 4 possible Rrs bands is used in the numerator for that particular row. *log_chl* is log transformed in-situ measured chlorophyll *a* concentration, originally in $mg\m ^{-3}$. *hplc_flag* is a binary variable indicating whether the measurement was obtained via HPLC or fluorometry. "-flag" are used during modeling for grouping data for hierarchical partial pooling models (see below.) {#data-table}
:::

@@

A sound Bayesian modeling principle is to develop several models with increasing levels of complexity as insight is progressively built. I largely follow this approach with the exception that the first model is built from a known formulation, a 4th order polynomial. All models use $log(MBR)$ as main input and $log(chl)$ as target, some use $MBR\_flag$ as additional input, the last model also uses $hlpc_flag$. The following are key components shared across models.

*Likelihood* - A truncated gaussian distribution with support constrained to -3.0 to 3.2 on the $log_{10}$ scale,corresponding to 0.001 - 1600 $mg\ m^{-3}$ to reflect both detectable realistic levels of marine surface chlorophyll.

@@

- No grouping or partial pooling  
- Constant noise term σ shared across all observations  
- Gamma prior placed on σ

>>>>

:::{#data-table}

|   | log_MBR              | MBR_flag   | log_chl              | hplc_flag   |
|----:|:-----------------------|:-----------|:-----------------------|:------------|
| 905 | -0.2545615131431814  | Rrs510     |  0.5837653682849998  | hplc        |
|  91 | -0.3641327253450254  | Rrs510     |  1.0511525224473812  | fluo        |
| 528 | -0.0737170739918758  | Rrs510     |  0.14363923527454328 | fluo        |
| 969 |  0.5813600497877703  | Rrs411     | -0.8272803386505035  | fluo        |
| 552 |  0.7134341855521412  | Rrs443     | -0.9318141382538383  | hplc        |

A random 5-row sample from the Pandas dataframe holding the data used in subsequent models. *log_MBR* is the log-transformed Max. Band Ratio (MBR) of sea surface reflectances, where the numerator is Max(Rrs411, Rrs443, Rrs489, Rrs510) and the denominator is computed from the sum of Rrs555 and Rrs670. *MBR_flag* indicates which of the 4 possible Rrs bands is used in the numerator for that particular row. *log_chl* is log transformed in-situ measured chlorophyll *a* concentration, originally in $mg\ m^{-3}$. *hplc_flag* is a binary variable indicating whether the measurement was obtained via HPLC or fluorometry. "-flag" are used during modeling for grouping data for hierarchical partial pooling models, explained further below. These groupings can be seen in [@fig-eda]

@@

::: {#fig-eda layout-ncol=2}

![](images/eda_numerator.png){#fig-eda-mbr}

![](images/eda_hplc.png){#fig-eda-hplc}

Two ways of grouping the data. [@fig-eda-mbr], max band ratio numerator; [@fig-eda-hplc], chlorophyll measurement method.
:::

@@

Several models of increasing levels were built.  All models use $log(MBR)$ as primary input and $log(chl)$ as target, some use $MBR\_flag$ as additional input, the last model also uses $hlpc_flag$. The following are key components shared across models.

*Likelihood* - A truncated gaussian distribution with support constrained to -3.0 to 3.2 on the $log_{10}$ scale,corresponding to $0.001 - 1600 mg\ m^{-3}$ to reflect both detectable realistic levels of marine surface chlorophyll.

@@

This model can be represented by the following set of probabilistic equations:

\begin{align*}
    \alpha &\sim \text{Normal}(0, 1) \\
    \beta_1 &\sim \text{Normal}(0, 1) \\
    \beta_2 &\sim \text{Normal}(0, 1) \\
    \beta_3 &\sim \text{Normal}(0, 1) \\
    \beta_4 &\sim \text{Normal}(0, 1) \\
    \mu &= \alpha + \beta_1 X + \beta_2 X^2 + \beta_3 X^3 + \beta_4 X^4 \\
    \sigma &\sim \text{Gamma}(2, 2) \\
    \text{likelihood} &\sim \text{TruncatedNormal}(\mu, \sigma, -3, 3.3)
\end{align*}

::: {#fig-oc6-struct}

![](images/oc6_structure.svg)
